---
title: "How to Replicate the Results in _Efficient Designs and Analysis of Two-Phase Studies with Longitudinal Binary Data_ by Di Gravio, Schildcrout, Tao (2023)"
author: "Chiara Di Gravio"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Replicate the Results in _Efficient Designs and Analysis of Two-Phase Studies with Longitudinal Binary Data_ by Di Gravio, Schildcrout, Tao (2023)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The vignette shows how to replicate the results presented in _"Efficient Designs and Analysis of Two-Phase Studies with Longitudinal Binary Data"_ by Di Gravio, Schildcrout and Tao (2023). 

The data used for the simulation studies summarized in Section 5 were generated using the following function:

```{r, eval = FALSE}
datGen <- function(betas, gammas, sigmas, N, nclust, ni, alpha0 = NULL, 
                   alpha1 = NULL,
                   xtype = "binary"){
  
  id       <-  sort(rep(seq(N), nclust))
  # binary X and Z
  if(xtype == "binary"){
     Z     <-  rbinom(n = N, size = 1, prob = 0.3)
     logod <-  alpha0 + alpha1*Z
     pX    <-  expit(logod)
     X     <- rbinom(n = N, size = 1, prob = pX)
  } else {
     Z    <-  rnorm(n = N)
     X    <-  rnorm(n = N, mean = -0.5 - alpha1*Z, sd = 1)
  }
  time <-  rep(c(0:(nclust-1)), N)
  Z    <- rep(Z, each = nclust)
  X    <- rep(X, each = nclust)
  dat  <- data.frame(id, X, Z, time)

  # generate y: transition model without Z*time interaction
  y    <- binaryMM::GenBinaryY(mean.formula= ~ X + time + Z + X:time,
                           t.formula = ~1, lv.formula = NULL, 
                           beta = betas, gamma = gammas,
                           sigma = sigmas, id = id, data = dat,
                           q = 20)
  dat <- data.frame(dat, Y = y)
  # generate balanced and incomplete data
  droptime <- rep(sample(seq(ni[1], ni[2]), N, replace=TRUE), each=ni[2]) - 1
  dat        <- dat[dat$time<=droptime,]
  dat
}
```

Note that the function requires users to load the `binaryMM` package in order to simulate the longitudinal binary outcome. There is no need to load the `dames` package to generate the data. 

**Simulation 1: Binary Expensive Exposure and Inexpensive Covariate**

```{r, eval = FALSE}
###### SMALL CLUSTER SIZE SCENARIO ######
# set up the parameters
params      <- c(-2.95, 0.25, 0.15, 1.5, 0.15, 2) 
N.sim       <- 2000
ni.sim      <- c(3, 6)
nclust.sim  <- max(ni.sim)
n_sieve     <- 10
betas       <- params[1:5]
gamma       <- params[6]
sigma       <- NULL
# a) correlated exposure and covariate
dat.1a    <- datGen(betas = betas, gammas = gamma, sigmas = sigma, 
                    ni = ni.sim, N = N.sim, nclust = nclust.sim,
                    alpha0 = -2.20, alpha1 = 2, xtype = "binary")

# b) independent exposure and covariate
dat.1b    <- datGen(betas = betas, gammas = gamma, sigmas = sigma, 
                    ni = ni.sim, N = N.sim, nclust = nclust.sim,
                    alpha0 = -1.35, alpha1 = 0, xtype = "binary")

```

**Simulation 2: Continuous Expensive Exposure and Inexpensive Covariate**

```{r, eval = FALSE}
###### SMALL CLUSTER SIZE SCENARIO ######
# set up the parameters
params      <- c(-2.5, 0.5, 0.15, 1.5, 0.15, 2)
N.sim       <- 1000
ni.sim      <- c(3, 6)
nclust.sim  <- max(ni.sim)
n.rand      <- 200

# a) correlated exposure and covariate
dat.2a    <- datGen(betas = betas, gammas = gamma, sigmas = sigma, 
                    ni = ni.sim, N = N.sim, nclust = nclust.sim, alpha1 = 0.2,
                    xtype = "continuous")
```

## Set-Up the Design

The paper compares none-some-all (NSA) designs and residual dependent sampling (RDS) designs. Both NSA and RDS designs can be implemented using functions in the `dames` package. Below is an example on how to perform NSA[10,480,10] and vR[600]:

```{r, eval = FALSE}
###### NSA DESIGN ######
nsa       <- dames::nsa_design(Y = "Y", id = "id", 
                               n.sample = c(10, 480, 10), data = dat.1a)
# set as missing the value of the expensive exposure X for subjects not sampled
dat.nsa   <- dat.1a
dat.nsa$X <- ifelse(dat.nsa$id %in% nsa$sampled.id, dat.nsa$X, NA)

###### RDS DESIGN ######
rds <- dames::res_design(Y = "Y", id = "id", 
                  mean.formula = Y ~ Z + time, t.formula = ~1, lv.formula = NULL,
                  n.sample = 600, statistic = "variance", data = dat.1a)
# set as missing the value of the expensive exposure X for subjects not sampled
dat.rds   <- dat.1a
dat.rds$X <- ifelse(dat.rds$id %in% rds$sampled.id, dat.rds$X, NA)
```

To implement a different NSA design, users can change the values of `n.sample` in the `nsa_design` call function. For instance, the NSA[0, 500, 0] design can be carried out by setting `n.sample = c(0, 500, 0)`. 

Similarly, one can implement a different RDS design by changing one input of the `res.design` function. Users can select `statistic = mean` for a mR[300, 300] design or `statistic = abs.mean` for a abs.mR[600] design.


## Running Estimator Code

Inference methods can be divided into two groups: methods that only include subjects with complete data on outcome, expensive exposure and inexpensive covariates (complete-case analyses), and methods that include all subjects regardless of whether they have information on the expensive exposure (full-cohort analyses). Below, we provide the code needed to replicate the results from the analyses presented in the paper.

**Ascertainment Corrected Maximum Likelihood (ACML) Estimator**

The estimated coefficients and standard errors can be estimated using the `mod.acml1` function. The function below refers to the NSA[10, 480, 10] design. If one is interested in a NSA[0,500,0] design, then the input of `samp.probs` should be changed into `c(0, 500/655, 0)`. The function call will be the same regardless on whether the expensive exposure and the inexpensive covariate are binary or continuous.

```{r, eval = FALSE}
# Simulation 1 binary covariates - small sample size
mod.acml1    <- dames::acml_mm(mean.formula = Y ~ X + time + Z + X:time,
                               t.formula = ~1, id = id, 
                               samp.probs = c(10/1325, 480/655, 10/25),
                               data = dat.nsa)
```

**Multiple Imputation**

The estimated coefficients and standard errors can be estimated using the `mod.acml1` function. The function below refers to the NSA[10, 480, 10] design. If one is interested in a NSA[0,500,0] design, then the input of `samp.probs` should be changed into `c(0, 500/655, 0)`. Note that the multiple imputation function cannot be used if the expensive exposure is continuous. 

```{r, eval = FALSE}
# define an indicator variable for sampling
dat.na$Sampled <- ifelse(dat.nsa$id %in% nsa$sampled.id, 1, 0)
# Simulation 1 - small sample size
mod.mi <- dames::mi_mm(mean.formula = Y ~ X + time + Z + X:time,
                 t.formula = ~1,
                 lv.formula = NULL, id = "id", X = "X",
                 sampled = "Sampled",
                 data = dat.nsa,
                 samp.probs = c(10/1325, 480/655, 10/25),
                 marg.exp.formula = X ~ Z, M = 5, method = "indirect")
```

**Sieve Maximum Likelihood Estimator (SMLE)**

```{r,eval = FALSE}
# Simulation 1 binary covariates - small sample size
# set up P(X|Z)
Bspline_Z <- cbind(as.numeric(dat.nsa$Z == 0), as.numeric(dat.nsa$Z == 1))
n_sieve             <- ncol(Bspline_Z)
colnames(Bspline_Z) <- paste("bs", 1:n_sieve, sep="")
exampledat          <- cbind(exampledat, Bspline_Z)
# estimate the parameters using SMLE
mod.smle <- smle_mm(mean.formula = Y ~ X + time + Z + X:time,
                    t.formula = ~1, lv.formula = NULL,
                    Y = "Y", X = "X", Z = "Z", Time = "time",
                    id = "id", Bspline = Bspline_Z, n_sieve = n_sieve, 
                    no_SE = FALSE,
                    data = dat.nsa)

# Simulation 2 continuous covariates - small sample size
# set up P(X|Z)
n_sieve               <- 10
Bspline_Z             <- bs(dat.nsa$Z, df = n_sieve, degree = 1,
                            Boundary.knots = range(dat.nsa$Z), intercept = TRUE)
colnames(Bspline_Z)   <- paste0("bs", 1:n_sieve)
dat.nsa                <- cbind(dat.nsa, Bspline_Z)
# estimate the parameters using SMLE
mod.smle <- smle_mm(mean.formula = Y ~ X + time + Z + X:time,
                    t.formula = ~1, lv.formula = NULL,
                    Y = "Y", X = "X", Z = "Z", Time = "time",
                    id = "id", Bspline = Bspline_Z, n_sieve = n_sieve, 
                    no_SE = FALSE,
                    data = dat.nsa)
```
